---
title: "p8105_hw3_zg2486"
output: github_document
name: Ziqi Guo
date: "2024-10-07"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
```



# Problem 1
1. Import dataset.
```{r}
library(p8105.datasets)
data("ny_noaa")

nrow(ny_noaa)
ncol(ny_noaa)
```
There are `r nrow(ny_noaa)` observations and `r ncol(ny_noaa)` variables in the dateset.The variables include `id`, `date`, `prcp`, `snow`, `snwd`, `tmax`, `tmin`. We can see lots of tmax and tmin data is missing and some prcp, snow, snwd data are missing as well.

2. Data cleaning.
```{r}
str(ny_noaa)

clean_df = ny_noaa %>% 
  janitor::clean_names() %>% 
  mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date),
    prcp = prcp / 10,
    snow = snow / 10,
    snwd = snwd / 10,
    tmax = as.numeric(tmax) / 10,
    tmin = as.numeric(tmin) / 10
  ) %>% 
  drop_na() 

clean_df %>% 
  count(snow, name = "n_obs") %>% 
  arrange(desc(n_obs))
```
The most commonly observed value is 0 cm.

3. Make a two-panel plot for average tmax.
```{r message = FALSE}
plot_df = clean_df %>% 
  mutate(
    year_month = lubridate::floor_date(date, unit = "month")) %>% 
  group_by(id, year_month) %>% 
  summarise(mean_tmax = mean(tmax))
  
ggplot(plot_df, aes(x = year_month, y = mean_tmax, color = id)) +
  geom_point(alpha = .5) +
  facet_grid(. ~ id)
  
```



# Problem 2
1. Import, clean, merge data sets.
```{r}
demo_df = read_csv("./nhanes_covar.csv",
                    na = c("NA", "", "."),
                    skip = 4,
                    show_col_types = FALSE) %>% 
  janitor::clean_names() %>% 
  filter(age >= 21) %>% 
  drop_na() %>% 
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    education = factor(education, levels = c(1, 2, 3), labels = c("Less than high school", "High school equivalent", "More than high school"))) 

accel_df = read_csv("./nhanes_accel.csv",
                    na = c("NA", "", "."), 
                    show_col_types = FALSE) %>% 
  janitor::clean_names() 

merge_df = left_join(demo_df, accel_df, by = c("seqn")) %>% 
  select(seqn, sex, age, bmi, education, everything())
```








