---
title: "p8105_hw3_zg2486"
output: github_document
name: Ziqi Guo
date: "2024-10-07"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
```



# Problem 1
1. Import dataset.
```{r}
library(p8105.datasets)
data("ny_noaa")

nrow(ny_noaa)
ncol(ny_noaa)
```
There are `r nrow(ny_noaa)` observations and `r ncol(ny_noaa)` variables in the dateset.The variables include `id`, `date`, `prcp`, `snow`, `snwd`, `tmax`, `tmin`. We can see lots of tmax and tmin data is missing and some prcp, snow, snwd data are missing as well.

2. Data cleaning.
```{r}
str(ny_noaa)

clean_df = ny_noaa %>% 
  janitor::clean_names() %>% 
  mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date),
    prcp = prcp / 10,
    snow = snow / 10,
    snwd = snwd / 10,
    tmax = as.numeric(tmax) / 10,
    tmin = as.numeric(tmin) / 10
  ) %>% 
  drop_na() 

clean_df %>% 
  count(snow, name = "n_obs") %>% 
  arrange(desc(n_obs))
```
The most commonly observed value is 0 cm.

3. Make a two-panel plot for average tmax.
```{r message = FALSE}
plot_df = clean_df %>%
  filter(month %in% c(1, 7)) %>% 
  mutate(
    month = case_when(
      month == 1 ~ "January",
      month == 7 ~ "July"
    )
  ) %>% 
  group_by(year, month, id) %>% 
  summarise(mean_tmax = mean(tmax))
  
ggplot(plot_df, aes(x = year, y = mean_tmax, color = id)) +
  geom_line(alpha = .5) +
  facet_grid(. ~ month) +
  theme(legend.position = 'None')
  
```



# Problem 2
1. Import, clean, merge data sets.
```{r}
demo_df = read_csv("./nhanes_covar.csv",
                    na = c("NA", "", "."),
                    skip = 4,
                    show_col_types = FALSE) %>% 
  janitor::clean_names() %>% 
  filter(age >= 21) %>% 
  drop_na() %>% 
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    education = factor(education, levels = c(1, 2, 3), labels = c("Less than high school", "High school equivalent", "More than high school"))) 

accel_df = read_csv("./nhanes_accel.csv",
                    na = c("NA", "", "."), 
                    show_col_types = FALSE) %>% 
  janitor::clean_names() 

merge_df = left_join(demo_df, accel_df, by = c("seqn")) %>% 
  select(seqn, sex, age, bmi, education, everything())
```

2. Produce a reader-friendly table and plot.
```{r message = FALSE}
merge_df %>% 
  group_by(sex, education) %>% 
  summarise(count = n()) %>% 
  pivot_wider(
     names_from = "sex",
    values_from = "count",
    values_fill = 0
  ) %>% 
  knitr::kable() 

ggplot(merge_df, aes(x = education, y = age, color = sex)) + 
  geom_boxplot() +
  theme(legend.position = "bottom")
```
From the boxplot, we can see similar median age of more than high school group and less than high school group, and the have similar distribution and range, but the median age of equivalent high school group for female is higher than male. Moreover, the overall age of have more than high school education level will be younger than other two education levels. 

3. Create total activity over the day and plot.
```{r message = FALSE}
total_activity = merge_df %>% 
  group_by(seqn, sex, age, bmi, education) %>% 
  summarise(total_mims = sum(across(min1:min1440)))

ggplot(total_activity, aes(x = age, y = total_mims, color = sex)) +
  geom_point(alpha =.5) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education) +
  theme(legend.position = "bottom")
```
The trends of total_activity for all female age in high school equivalent and more than higher school groups are higher than male. But in less than high school group, the trend of male surpass the trend of female around 41-42 years old. And the overall trend of male and female in each education level is decreasing with age increasing. 

4. Make a three-panel plot.
```{r message = FALSE}
activity_time = merge_df %>% 
  pivot_longer(
    min1:min1440,
    names_to = "minute",
    names_prefix = "min",
    values_to = "mims"
  ) %>% 
  mutate(minute = as.numeric(minute)) %>% 
  group_by(sex, education, minute) %>% 
  mutate(mean_mims = mean(mims)) 

ggplot(activity_time, aes(x = minute, y = mean_mims, color = sex)) +
  geom_line(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education) +
  theme(legend.position = "bottom")
```
Female and male have a similar trend in 24-hour mean_mims activity time courses for each education level. For less than high school level, female have lower mean_mims than male when minute < 550, and female have higher mean_mims than female when minute around 550 - 750. For high school equivalent level, female overall have a higher mean_mims than male, especially when minute is around 500 -1000. For more than high school level, female overall have a higher mean_mims than male as well, especially when minute is around 
500 -1250. 



# Problem 3
1. Import, clean, and tidy data.
```{r message = FALSE}
jan_2020_df = 
  read_csv("./citibike/Jan 2020 Citi.csv",
        na = c("NA", "", ".")) %>% 
  janitor::clean_names() %>% 
  mutate(month = "january") %>% 
  mutate(year = "2020") %>% 
  mutate(date = "january 2020")

jan_2024_df = 
  read_csv("./citibike/Jan 2024 Citi.csv",
            na = c("NA", "", ".")) %>% 
  janitor::clean_names() %>% 
  mutate(month = "january") %>% 
  mutate(year = "2024") %>% 
  mutate(date = "january 2024")

july_2020_df = 
  read.csv("./citibike/July 2020 Citi.csv",
            na = c("NA", "", ".")) %>% 
  janitor::clean_names() %>% 
  mutate(month = "july") %>% 
  mutate(year = "2020") %>% 
  mutate(date = "july 2020")

july_2024_df = 
  read.csv("./citibike/July 2024 Citi.csv",
            na = c("NA", "", ".")) %>% 
  janitor::clean_names() %>% 
  mutate(month = "july") %>% 
  mutate(year= "2024") %>% 
  mutate(date = "july 2024")

final_df = bind_rows(list(jan_2020_df, jan_2024_df, july_2020_df, july_2024_df)) 
```

2. Create a reader-friendly table.
```{r message = FALSE}
final_df %>% 
  group_by(date, member_casual) %>% 
  summarise(count = n()) %>% 
  pivot_wider(
    names_from = "member_casual",
    values_from = "count",
  ) %>%
  knitr::kable()
```
From the table, we can see overall most riders are member and members have more riding than casual riders. There were over 45000 riders in July 2024.

3. Make a table showing the 5 most popular starting stations for July 2024.
```{r}
july_2024_df %>% 
  group_by(start_station_name) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count)) %>% 
  slice_head(n =5) %>% 
  knitr::kable()
```

4. Make a plot to investigate the effects of day of the week, month, and year on median ride duration.
```{r message = FALSE}
final_df %>% 
  mutate(
    weekdays = factor(weekdays, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
  ) %>% 
  group_by(weekdays, month, year) %>%
  summarise(median_duration = median(duration)) %>% 
  ggplot(aes(x = weekdays, y = median_duration, color = interaction(year, month))) +
  geom_line(aes(group = interaction(year, month))) + 
  facet_grid(year ~ month) +
  theme(legend.position = "bottom")
```
Comparing each day of the week, we can see generally there were higher median ride duration on the weekend than weekdays, besides Sunday on January 2024. We can also see no matter of which year, the median ride duration in July were higher than January. Moreover, the median ride duration were higher in 2020 compared to 2024 regardless of month.

5. Make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration for 2024 data.
```{r}
final_df %>% 
  filter(year == "2024") %>% 
  ggplot(aes(x = member_casual, y = duration, color = rideable_type)) +
  geom_boxplot () +
  facet_grid(. ~ month) +
  theme(legend.position = "bottom")
```






